```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(stringr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(readxl)

defense = read_excel('cyclonesFootball2019.xlsx', sheet='Defensive')
offense = read_excel('cyclonesFootball2019.xlsx', sheet='Offensive')
bio = read_excel('cyclonesFootball2019.xlsx', sheet='Biography')
```

---
title: Team Work file
author: Aaron Jencks, Eli Musgrove
date: 3/19/2020
output: html_document
---

## Part 1: Cleaning

1. The player names and opponent names are factors

```{r}
# Create new dataframes to be cleand
defClean = data.frame(defense)
offClean = data.frame(offense)
bioClean = data.frame(bio)

# Factorize player names
total_names_list = unique(c(defense$Name, offense$Name, bio$Name))
defClean$Name = factor(defense$Name, levels=total_names_list)
offClean$Name = factor(offense$Name, levels=total_names_list)
bioClean$Name = factor(bio$Name, levels=total_names_list)

# Factorize opponent names
total_opponent_names_list = unique(c(defense$Opponent_Opponent, offense$Opponent_Opponent))
defClean$Opponent_Opponent = factor(defense$Opponent_Opponent, levels=total_opponent_names_list)
offClean$Opponent_Opponent = factor(offense$Opponent_Opponent, levels=total_opponent_names_list)
```

2. All offensive and defensive statistics are numerical. The Weight column in biography is also numeric.

```{r}
defClean = defClean %>% mutate_at(colnames(defClean)[3:11], as.numeric)
offClean = offClean %>% mutate_at(colnames(offClean)[3:12], as.numeric)
bioClean$Weight = as.numeric(bioClean$Weight)
```

3. Change the Height column in biography into numeric.

```{r}
bioClean = bioClean %>% mutate(height_feet = (function(element) {
  
  # You'll need the stringr package for these methods, 
  # they are used to splice the feet and inches parts of the Height
  index = (element %>% str_locate("-"))[,1]
  result = as.numeric(element %>% str_sub(rep(1, length(index)), index - 1))
  return(result)

})(Height),

Height = (function(elements) {

  # Converts feet into inches, and then adds the inches
  index = (elements %>% str_locate("-"))[,1]
	feet = as.numeric(elements %>% str_sub(1, index - 1)) * 12
	inches = as.numeric(elements %>% str_sub(index + 1, rep(length(elements), length(elements))))
	result = feet + inches
	return(result)

})(Height))
```


```{r}
# Final clean dataframes
str(defClean)
str(offClean)
str(bioClean)
```

## Part 2: Tidying

### Part 1

1. Reformat the defClean data frame into tidy format using `pivot_longer`. The type of statistic (Tackles_Solo, Tackles_ASST, etc) is added as a new key column named Statistic.

```{r}
defClean = defClean %>%
  pivot_longer(Tackles_Solo:Pass_PB, names_to='Statistic', values_to='Value')
```

2. Create a histogram for each defensive statistic. Use `facet_wrap` to draw all histograms in a single plot. What defensive skills are rare?

##### The defensive skills 'Turnover_FF', 'Turnover_FR', and 'Turnover_INT' are rare with most values close to 0. This is because turnovers are pretty rare in the course of a game while tackling would occur more often.

```{r}
ggplot(defClean, aes(x=Value)) + geom_histogram() + facet_wrap(~Statistic) + stat_bin(bins=10)
```

3. Did ISU have better defense against West Virginia or Kansas State? Answer this question by creating a scatterplot with x-axis being the number of solo tackles against West Virginia, and y-axis being that against Kansas State. A large number of solo tackles indicates a good defense.

##### The data doesn't point to a strong conclusion- a large number of players had 0 tackles against either opponent, but there are a few outliers that skew the regression line to have a smaller slope. Although the regression line doesn't indicate a strong conclusion, it shows there is a slight favor of Iowa State's defense towards being stronger against West Virginia. If there were more data points a stronger conclusion could be drawn, but not enough tackles can be made in the given sample to have a strong definitive conclusion.
```{r}
# Prep dataframe
defComp = defClean %>% filter(Statistic == 'Tackles_Solo') %>% filter(Opponent_Opponent %in% c('West Virginia', 'Kansas State')) %>%       pivot_wider(names_from=Opponent_Opponent, values_from=Value)

colnames(defComp) = c('Name', 'Statistic', 'West_Virginia', 'Kansas_State')
ggplot(defComp, aes(x=West_Virginia, y=Kansas_State)) + geom_smooth(method='lm') + geom_point(position='jitter') + ggtitle('Iowa State University Solo Tackles')
```

### Part 2

1. Separate the Hometown column into two columns, namely the city and the state. Print the first few rows of these two columns.

```{r}
bioClean = bioClean %>% separate(Hometown, c('City', 'State'), sep=',', remove=FALSE)
head(bioClean['City'])
head(bioClean['State'])
```

2. How many players are there from each state?

```{r}
fromStates = bioClean %>% group_by(State) %>% tally()
print(fromStates, n=Inf)
```

## Part 3: Joining data frames

1. Find where the offensive players come from by joining the offensive data and player biography data appropriately.

```{r}
# Use full_join, only because it wasn't specified to get rid of N/A values
hometowns = full_join(offClean, bioClean, by="Name") %>% select(Name, Hometown) %>% distinct()
head(hometowns)
```

2. Is there a relationship between the weight of a player and the receiving yards (`Receiving_YDS`)?

##### There is a small linear correlation between the `Weight` and the `mean_yds`. The black points are the raw data, and the colored points are the average. If you don't assume that the least weighing student is an outliar, then there is a small exponential relationship, with the valley right around 175.

```{r}
# Use inner_join this time, because we aren't interested in N/A values
weight_yards = inner_join(bioClean, offClean, by="Name") %>% 
	       select(Name, Weight, Receiving_YDS) %>% 
	       group_by(Name) %>% 
	       mutate(mean_yds = mean(Receiving_YDS))

ggplot(distinct(weight_yards, Name, .keep_all=TRUE), aes(x=Weight, y=mean_yds, color=Name)) + ylab("Receiving Yards") +
	geom_point(data=weight_yards, aes(x=Weight, y=Receiving_YDS), show.legend=FALSE, inherit.aes=FALSE) + 
	geom_point(size=3)
```

3. How does the performance of quarterback Brock Purdy relate to the performance for the ISU team?

##### As a quarterback, Brock's performance should be compared to the other offensive players. One of the most important statistics for quarterbacks is the passing yards. Compared to the other players with non-null passing yards, Brock is far superior in comparision.

```{r}
# Join bio and offensive tables and select passing_yds
perf = full_join(bioClean, offClean, by='Name') %>% group_by(Name) %>% select(Name, Passing_YDS)
perf = na.omit(perf)

ggplot(perf, aes(y=Passing_YDS)) + geom_boxplot() + facet_wrap(~Name)
```

4. Explore more. You may in addition explore the 2018 data (see slides). Raise a question that can be answered only by looking at two or more data frames. Then answer that question.

##### Question: Among the offensive players, which state produced the most players? Which states produce the most defensive players?

##### The states Florida and Oklahoma are tied with 2 players from each state. Since offensive positions are very important, coaches are more likely to look outside of Iowa to draft the best talent they can find.

```{r}
offPlay = inner_join(bioClean, offClean, by='Name') %>% group_by(State) %>% distinct(Name) %>% tally(sort=TRUE)
head(offPlay, 3)
```

##### Iowa produced the most defensive players. These positions are the least likely to have money spent to recruit outside the state for. It also seems like there are a lot of players overall that come from Florida to play in Iowa since there are 6 overall players between offensive and defensive from Florida.

```{r}
defPlay = inner_join(bioClean, defClean, by='Name') %>% group_by(State) %>% distinct(Name) %>% tally(sort=TRUE)
head(defPlay, 2)
```