```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(stringr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(readxl)

defense = read_excel('cyclonesFootball2019.xlsx', sheet='Defensive')
offense = read_excel('cyclonesFootball2019.xlsx', sheet='Offensive')
bio = read_excel('cyclonesFootball2019.xlsx', sheet='Biography')
```

---
title: Team Work file
author: Aaron Jencks, Eli Musgrove
date: 3/19/2020
output: html_document
---

## Part 1: Cleaning

1. The player names and opponent names are factors

```{r}
# Create new dataframes to be cleand
defClean = data.frame(defense)
offClean = data.frame(offense)
bioClean = data.frame(bio)

# Factorize player names
total_names_list = unique(c(defense$Name, offense$Name, bio$Name))
defClean$Name = factor(defense$Name, levels=total_names_list)
offClean$Name = factor(offense$Name, levels=total_names_list)
bioClean$Name = factor(bio$Name, levels=total_names_list)

# Factorize opponent names
total_opponent_names_list = unique(c(defense$Opponent_Opponent, offense$Opponent_Opponent))
defClean$Opponent_Opponent = factor(defense$Opponent_Opponent, levels=total_opponent_names_list)
offClean$Opponent_Opponent = factor(offense$Opponent_Opponent, levels=total_opponent_names_list)
```

2. All offensive and defensive statistics are numerical. The Weight column in biography is also numeric.

```{r}
defClean = defClean %>% mutate_at(colnames(defClean)[3:11], as.numeric)
offClean = offClean %>% mutate_at(colnames(offClean)[3:12], as.numeric)
bioClean$Weight = as.numeric(bioClean$Weight)
```

3. Change the Height column in biography into numeric.

```{r}
bioClean = bioClean %>% mutate(height_feet = (function(element) {
  
  # You'll need the stringr package for these methods, 
  # they are used to splice the feet and inches parts of the Height
  index = (element %>% str_locate("-"))[,1]
  result = as.numeric(element %>% str_sub(rep(1, length(index)), index - 1))
  return(result)

})(Height),

Height = (function(elements) {

  # Converts feet into inches, and then adds the inches
  index = (elements %>% str_locate("-"))[,1]
	feet = as.numeric(elements %>% str_sub(1, index - 1)) * 12
	inches = as.numeric(elements %>% str_sub(index + 1, rep(length(elements), length(elements))))
	result = feet + inches
	return(result)

})(Height))
```


```{r}
# Final clean dataframes
str(defClean)
str(offClean)
str(bioClean)
```

## Part 2: Tidying

### Part 1

1. Reformat the defClean data frame into tidy format using `pivot_longer`. The type of statistic (Tackles_Solo, Tackles_ASST, etc) is added as a new key column named Statistic.

```{r}
defClean = defClean %>%
  pivot_longer(Tackles_Solo:Pass_PB, names_to='Statistic', values_to='Value')
```

2. Create a histogram for each defensive statistic. Use `facet_wrap` to draw all histograms in a single plot. What defensive skills are rare?

##### The defensive skills 'Turnover_FF', 'Turnover_FR', and 'Turnover_INT' are rare with most values close to 0. This is because turnovers are pretty rare in the course of a game while tackling would occur more often.

```{r}
ggplot(defClean, aes(x=Value)) + geom_histogram() + facet_wrap(~Statistic) + stat_bin(bins=10)
```

3. Did ISU have better defense against West Virginia or Kansas State? Answer this question by creating a scatterplot with x-axis being the number of solo tackles against West Virginia, and y-axis being that against Kansas State. A large number of solo tackles indicates a good defense.

##### The data doesn't point to a strong conclusion- a large number of players had 0 tackles against either opponent, but there are a few outliers that skew the regression line to have a smaller slope. Although the regression line doesn't indicate a strong conclusion, it shows there is a slight favor of Iowa State's defense towards being stronger against West Virginia. If there were more data points a stronger conclusion could be drawn, but not enough tackles can be made in the given sample to have a strong definitive conclusion.
```{r}
# Prep dataframe
defComp = defClean %>% filter(Statistic == 'Tackles_Solo') %>% filter(Opponent_Opponent %in% c('West Virginia', 'Kansas State')) %>%       pivot_wider(names_from=Opponent_Opponent, values_from=Value)

colnames(defComp) = c('Name', 'Statistic', 'West_Virginia', 'Kansas_State')
ggplot(defComp, aes(x=West_Virginia, y=Kansas_State)) + geom_smooth(method='lm') + geom_point(position='jitter') + ggtitle('Iowa State University Solo Tackles')
```

### Part 2

1. Separate the Hometown column into two columns, namely the city and the state. Print the first few rows of these two columns.

```{r}
bioClean = bioClean %>% separate(Hometown, c('City', 'State'), sep=',')
head(bioClean['City'])
head(bioClean['State'])
```

2. How many players are there from each state?

```{r}
fromStates = bioClean %>% group_by(State) %>% tally()
print(fromStates, n=Inf)
```

## Part 3: Joining data frames

1. Find where do the offensive players come from by joining the offensive data and player biography data appropriately.

	```{r}

	# Use full_join, only because it wasn't specified to get rid of N/A values
	hometowns = full_join(offense, bio, by="Name") %>% select(Name, Hometown) %>% distinct()
	head(hometowns)

	```

2. Is there a relationship between the weight of a player and the receiving yarsd (`Receiving_YDS`)?

	```{r}

	# Use inner_join this time, because we aren't interested in N/A values
	weight_yards = inner_join(bio, offense, by="Name") %>% 
		       select(Name, Weight, Receiving_YDS) %>% 
		       group_by(Name) %>% 
		       mutate(mean_yds = mean(Receiving_YDS))

	ggplot(distinct(weight_yards, Name, .keep_all=TRUE), aes(x=Weight, y=mean_yds, color=Name)) + ylab("Receiving Yards") +
		geom_point(data=weight_yards, aes(x=Weight, y=Receiving_YDS), show.legend=FALSE, inherit.aes=FALSE) + 
		geom_point(size=3)

	```

	There is a small linear correlation between the `Weight` and the `mean_yds`. The black points are the raw data, and the colored points are the average. If you don't assume that the least weighing student is an outliar, then there is a small exponential relationship, with the valley right around 175.

3. How do the performance of quarterback Brock Purdy relate to the performance fo the ISU team?
4. Explore more. You may in addition explore the 2018 data (see slides). Raise a question that can be answered only by looking at two or more data frames. Then answer that question.
